BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-04-24 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
edits from printed

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual generation of herbarium specimens is essential for current and future research on evolution and ecology. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau provides administrative jurisdictional information and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as extraneous site names. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by group experts [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. Even if a collector navigates all these hurdles, the time allocated to each step is quite large.

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
It rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage, which we have supplemented with over *XXXX* valleys allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience, it is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, it does not include data cleaning functionalities beyond verifying nations of collection. 
While some sources indicate that it can only be used in Microsoft, we expect it to be accessed on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

BarnebyLives was named by plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William and Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until he passed in 2000 [@welsh2001rupert]. 
Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert]. 
Most germanely, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].

@@

BarnebyLives was iteratively developed based on data submitted by approximately 20 seasonal field botany teams over two years. 
Essentially, continual updates were made as the developers became aware of the idiosyncrasies of collection notes and data entry.

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available Rstudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 90m resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a terminal.
Unfortunately, we did not find Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system. 
BarnebyLives was iteratively developed based on the data submitted by approximately 20 seasonal field botanists. 
Essentially, additions to the code have been continually made, as developers were exposed to more idiosyncrasies of collection notes and data entry.
Several commands in BarnebyLives require output from previous functions, and a workflow that satisfies these requirements is presented in Figure 1.

@@

The taxonomic module has four required functions and one optional function. 
*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World database filtered to the local continents or a user-specified backbone. 
*spell_check_family* performs a spell check on the family entered for each scientific name. author_check ensures that the authors are entered in a valid format, for example, the correct standard abbreviations are used. 
*associates_check* performs a spell check on all associated species using the local taxonomic database.
*powo_searcher* can be used in tandem with the functions spell_check_family and author-check, but we use it in lieu of them to search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species. No API key or registration is required to use powo_searcher.

The formatting module has three functions. The first two we will detail are technically optional; however, they are run locally and so quickly that there is no reason to skip them. 
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.

The manual review process technically only has one function that is optional and may be executed during the spatial process (after coords2sf), but the importance of manual review is important enough to warrant explicit mention. 
*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/). 
Notably, many of the flags that BarnebyLives generates will be placed into columns with obviously flagged names and can be manually reviewed by the analyst, and many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.   "

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites. 
BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.

Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec. 
The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` **sheets** to a single Portable Document Format (PDF). 
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

## Family Level Results
# these species were found by POW, and according to POW, the species are in the f
# following families. 
family_incongruence <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Family != 'NOT FOUND') %>% 
  filter(POW_Family != Family) %>% 
  select(Collection_number, Scientific_name, Family, POW_Family) %>% 
  mutate(Situation = 
    case_when(
      Family %in% c('Hydrophyllaceae', 'Namaceae') ~ 'Preferred', 
      Family %in% c('Athyriaceae') ~ 'Outdated',
      Family %in% c('Saxifragaceae', 'Plantaginaceae') ~ 'Submitter_Incorrect', 
      Family %in% c('Ranunculaceae', 'Ericaceae') ~ 'Internal_Error', 
      .default = 'Typo'
    ))

@@

epithet_diff <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(Epithet != POW_Epithet, !Scientific_name %in% genus_mismatch$Scientific_name) %>% 
  select(Collection_number, Scientific_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
  distinct(Scientific_name, .keep_all = T) %>%  # do we record repeat collections ?
  mutate(Situation = 
           case_when(
             Collection_number %in% c('1207') ~ 'User Error', 
             Collection_number %in% c(
               '1217', '1251', '1283', '1311', '1391', '1415', '1435', '1450', '1556','1576', 
               '1594', '1624', '1627', '1745', '2043', '2108', '2109', '2120', '2127', '2130', '2135', 
               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2370', '2379', 
               '2393', '2480', '2511', '2561', '2573', '2648', '2660', 
               '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
             ~ 'Spelling',
             Collection_number %in% c(
               '1337', '1421', '1588', '1619','2415', '2445', '2777', '1337', '2004', '2086', '2343', '2412', '2487', '2748', '2652',
               '2805', '2194', '2040', '2039', '2403',  '2076') 
             ~ 'Outdated', 
             Collection_number %in% c('2644') ~ 'Incorrect',
             Collection_number %in% c(
               '1207', '2326', '2038', '1424', '2618', '2622', '2665', '2711', '2713') 
             ~ 'Internal_Error',
             Collection_number %in% c('2110', '2252', '2328', '2540') ~ 'Preferred'
           )
  )

rm(Family_in, Family_out)

## check spell checks on both vegetation and associates columns

veg_diff <- data_p %>% 
  sf::st_drop_geometry() %>% 
  select(Collection_number, Vegetation, SpellCk.Vegetation) %>% 
  filter(Vegetation != SpellCk.Vegetation) %>% 
  distinct(Vegetation, SpellCk.Vegetation, .keep_all = TRUE) %>%

  # grouped sites - these sites are essentially replicates of each other, I have just
  # remove a focal taxon from them. I feel they are pseudoreplicates so they will be discarded

  filter(!Collection_number %in%
   c('2058', '2107', '2110', '2112', '2117', '2220', '2221', '2407', '2428', '2429', 
     '2430', '2432', '2540', '2541', '2560', '1407') 
  ) %>% 
  mutate(Situation = 
           case_when(
             Collection_number %in% c(  
               '1207', '1286', '1596', '1598', '2004', '2033', '2034', '2133', 
               '2175', '2402', '2407', '2427', '2760' 
               ) ~ 'User Error',
             Collection_number %in% c(
               '1214', '1218', '1220', '1227', '1228', '1250', '1257', '1258', '1270', 
               '1273', '1287', '1292', '1293', '1302', '1341', '1343', '1383', 
               '1384', '1390', '1401', '1455', '1557', '1558', '1565', '1596', 
               '2000', '2010', '2012', '2014', '2032', '2049', '2054', '2067', '2105', 
               '2116', '2163', '2219', '2224', '2230', '2395', '2406', '2454', '2528', 
               '2539', '2562', '2566', '2571', '2572', '2622', '2625', '2656', 
               '2664', '2668', '2745', '2759', '2787'
             )
             ~ 'Spelling',
             Collection_number %in% c(
                '1278', '1339', '1405'
               ) ~ 'Incorrect', # that is the program returns the wrong taxon
             Collection_number %in% c(
               '2527' 
             ) 
             ~ 'Internal_Error',
           )
  )
```

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. 
While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied. 
At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.

@@

<div id="refs"></div>

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual generation of herbarium specimens is essential for current and future research on ecology, evolution, and conservation. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau for administrative jurisdictional data, and other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions and locality information. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. Even if a collector navigates all these hurdles, the time allocated to each step is quite large.

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
It rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, it does not include data cleaning functionalities beyond verifying nations of collection. 
While some sources indicate that it can only be used in Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William and Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until he passed in 2000 [@welsh2001rupert]. 
Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert]. 
Most germanely, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].

@@

BarnebyLives was iteratively developed based on data submitted by approximately 20 seasonal field botany teams over two years. 
Essentially, continual updates were made as the developers became aware of the idiosyncrasies of collection notes and data entry.  
Several commands in BarnebyLives require output from previous functions, and a workflow that satisfies these requirements is presented in Figure 1.

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available RStudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system.

@@

The taxonomic module has four required functions and one optional function. 
*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World database filtered to the local continents or a user-specified backbone.  
*spell_check_family* performs a spell check on the family entered for each scientific name.  
*author_check* ensures that the authors are entered in a valid format, for example, the correct standard abbreviations are used.  
*associates_check* performs a spell check on all associated species using the local taxonomic database. 
*powo_searcher* can be used in tandem with the functions *spell_check_family* and *author_check*, but we use it in lieu of them to search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species. No API key or registration is required to use *powo_searcher*.

The formatting module has three functions. The first two we will detail are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.

The manual review process technically only has one function that is optional and may be executed during the spatial process (after *coords2sf*), but the importance of manual review is important enough to warrant explicit mention. 
*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/). 
Notably, many of the flags that BarnebyLives generates will be placed into columns with obviously flagged names and can be manually reviewed by the analyst, and many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.

Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec. 
The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the 182 sheets to a single Portable Document Format (PDF). 
The total label generation run time for processing these `r n_spp` collections was `r minutemen(sum(time_trials$elapsed))` minutes. 
In total the `r n_spp` collections, which underwent all processing steps, took `minutemen(sum(time_trials$elapsed, time_label_gen, as.numeric(time_final)))` minutes to process.

@@

## Family Level Results
# these species were found by POW, and according to POW, the species are in the f
# following families. 
family_incongruence <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Family != 'NOT FOUND') %>% 
  filter(POW_Family != Family) %>% 
  select(Collection_number, Scientific_name, Family, POW_Family) %>% 
  mutate(Situation = 
    case_when(
      Family %in% c('Hydrophyllaceae', 'Namaceae') ~ 'Preferred', 
      Collection_number %in% c('1594', '2375', '2547', '2563', '2729') ~'Autofilled', 
      Collection_number %in% c('1576', '2229', '2376', '2401', '2603', '2688') ~ 'Submitter_Incorrect', 
      Collection_number %in% c('2377') ~ 'Outdated', 
      Collection_number %in% c('2284', '2435', '2517', '2622', '2705', '2725', '2726') ~ 'Internal_Error', 
      .default = 'Typo'
    ))

@@

epithet_diff <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(Epithet != POW_Epithet, !Scientific_name %in% genus_mismatch$Scientific_name) %>% 
  select(Collection_number, Scientific_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
  distinct(Scientific_name, .keep_all = T) %>%  # do we record repeat collections ?
  mutate(Situation = 
           case_when(
             Collection_number %in% c('1207') ~ 'User Error', 
             Collection_number %in% c(
               '1217', '1251', '1283', '1311', '1337', '1391', '1415', '1421', '1435', '1450', '1576', 
               '1594', '1624', '1627', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130', '2135', 
               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2343', '2370', '2379', 
               '2393', '2415', '2480', '2511', '2561', '2573', '2648', '2660', 
               '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
             ~ 'Spelling',
             Collection_number %in% c(
               '1556', '1588', '1619', '2445', '2777', '1337', '2004', '2412', '2487', '2748', '2652',
               '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
             ~ 'Outdated', 
             Collection_number %in% c('2644') ~ 'Incorrect',
             Collection_number %in% c(
               '1207', '2326', '2038', '1424', '2618', '2622', '2665', '2711', '2713') 
             ~ 'Internal_Error',
             Collection_number %in% c('2110', '2540') ~ 'Preferred'
           )
  )

rm(Family_in, Family_out)
```

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. 
While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)`records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied. 
At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.

@@

\clearpage
<div id="refs"></div>
